Title Text

### 2014-04-06 paper.tex
Flesh out the comparison between MoA and Pull arrays

\authorinfo{Josef Svenningsson}
           {Department of Computer Science of Engineering\\Chalmers University of Technology}
           {josef.svenningsson@chalmers.se}
\authorinfo{Lenore M. Mullin}
           {Emeritus Professor\\Department of Computer Science\\
University at Albany, SUNY}
           {lenore@cs.albany.edu}

@@

Computing with arrays has been a constantly important paradigm in the
history of computing. For this reason we have seen a plethora of
formalisms and languages being developed to aid theoriticians and
developers to produce more suitable abstractions and faster programs.
Examples include Fortran, APL and Matlab but the list goes on and on.
Having many different options to choose from can be a good thing but
it can also lead to a balkanization of the community. When researcher
dedicated to different formalisms can no longer talk to each other it
inhibits collaborations and limits the impact of innovatation.

@@

\section{Overview of MoA}
\label{sec:moa}

\input{intromoa}

\section{Overview of Pull arrays}
\label{sec:pull}

@@

Arrays are represented as functions from index to element. We refer to
this function as an \emph{index function}. Additionally arrays also
has a length.

@@

@@

* Only one element type in MoA
* Not much in the way of types in MoA

* Shapes are arrays in MoA whereas they are a different type with Pull arrays

\subsection{\(\psi\)-reductions and fusion}
\label{sec:normalization}

The similarity between MoA and Pull arrays is most clear when
considering Operatonal Normal Form (ONF). Recall that ONF means that

@@

\begin{tabular}{|@{}l|l@{}|}
\hline
MoA & Feldspar
\\
\hline
δ & dim . extent
\\
ρ & extent
\\
ιξⁿ & -- generalized enumFromTo
\\
ψ & index
\\
rav & -- flattens to a one-dimensional array
\\
γ & toIndex
\\
γ' & fromIndex
\\
$s \: \hat{ρ} \: ξ$ & reshape
\\
π x & product
\\
τ & size
\\
++ & ++
\\
ξ₁ f ξ₂ & zipWith f ξ₁ ξ₁
\\
σ f ξ & map f ξ
\\
ξ f σ & map f ξ
\\
$\bigtriangleup$ & take
\\
$\bigtriangledown$ & drop
\\
$op^{red}$ & fold
\\
Φ & reverse
\\
Θ & rotate
\\
$\varobslash$ & -- generalized transpose
\\
$f Ω_d ξ$ & -- Apply $f$ along the dimension $d$ of $ξ$
\\
\hline
\end{tabular}

>>>>

\authorinfo{Josef Svenningsson}
           {Department of Computer Science of Engineering\\Chalmers University of Technology}
           {josef.svenningsson@chalmers.se}
\authorinfo{Lenore M. Mullin}
           {Emeritus Professor\\Department of Computer Science\\
University at Albany, SUNY}
           {lenore@albany.edu}

@@

Computing with arrays has been a constantly important paradigm in the
history of computing. For this reason we have seen a plethora of
formalisms and languages being developed to aid theoriticians and
developers to produce more suitable abstractions and faster programs.
Examples include Fortran, APL and Matlab but the list goes on and on.
Having many different options to choose from can be a good thing but
it can also lead to a balkanization of the community. When researchers
dedicated to different formalisms can no longer talk to each other,  it
inhibits collaborations and limits the impact of innovation.

@@

\section{Overview of MoA}
\label{sec:moa}
\input{intromoa}
\section{Overview of Pull arrays}
\label{sec:pull}

@@

Arrays are represented as functions from index to element. We refer to
this function as an \emph{index function}. Additionally arrays also
have a length.

@@

In this section we relate MoA and Pull arrays.

An initial observation is that MoA and Pull arrays where developed for
different purposes and that can be seen already by looking at the
notation. Mathematics of arrays have first and foremost been used for
calculating array programs by hand. This is reflected in the short
names and succinct notation which works very well when doing
pencil-and-paper proofs.

Pull arrays on the other hand where conveived in the context of a
typed functional language. The notation is completely inherited from
the syntax of Haskell \cite{marlow2010haskell}.

@@

When it comes to types MoA keeps things very simple. There are two
types: arrays and scalars. Furthermore, scalars and singleton arrays
are identified.

Pull arrays inherits the type discipline from the language they are
implemented in. When used in Haskell it means that they have a rich
type system to utilize. First and foremost, Pull arrays are
parametric, meaning that they can carry elements of any type. In
particular the elements can be Pull arrays as another way of
representing higher dimensional arrays. This representation is seldom
used as it has poor performance.

Shapes in Pull arrays are represented by a completely different type,
whereas in MoA shapes are simply arrays. Representing shapes as arrays
is challenging in a typed language such as Haskell because it leads to
a form of infinite recursion and while there are a number of way to
break the recursion we have yet to find a way that does so and retains
the efficiency and simplicity of Pull arrays.

\subsection{DNF, \(\psi\)-reductions and fusion}
\label{sec:normalization}

The correspondence between MoA and Pull arrays is most clear when
considering Denotaional Normal Form (DNF), \(\psi\)-reductions and
fusion. Both formalisms define operations on arrays by how to index
into the array. In MoA, the indexing function \(\psi\) is used,
whereas for Pull arrays operations are defined as functions from index
to the element at that index. These two forms are equivalent.

Pull arrays are always kept on the form of an indexing function and a
length. This representation is preserved for all operations. In MoA,
arrays are treated as abstract during derivations. After calculations,
\(\psi\)-reductions are performed to bring the expression into
Denotational Normal Form. DNF is equivalent to Pull arrays, it shows
how to compute an element from an array directly from an index.

The process of applying \(\psi\)-reductions corresponds to performing
fusion on Pull arrays. In both cases operations are inlined until
there is a single expression for computing an element from an array
given an index.

Pull arrays use the function \verb!store! to prevent fusion and store
an array in memory. MoA deals with memory differently. All arrays
which are supposed to be allocated in memory are given a separate
definition.

The last step in calculating a program in MoA is to derive an
Operational Normal Form which is explicit about how the array is laid
out in memory. Pull arrays do not have anything corresponding to
ONF. Instead, the decision of how the array is laid out is done by the
compiler, which uses a pre-defined convention.

@@

\begin{tabular}{|@{}l|l@{}|}
\hline
MoA & Feldspar
\\
\hline
δ & \verb!dim . extent!
\\
ρ & \verb!extent!
\\
ιξⁿ & generalized \verb!enumFromTo!
\\
ψ & generalized \verb!index!
\\
rav & -- flattens to a one-dimensional array
\\
γ & \verb!toIndex!
\\
γ' & \verb!fromIndex!
\\
$s \: \hat{ρ} \: ξ$ & \verb!reshape!
\\
π x & \verb!product!
\\
τ & \verb!size!
\\
++ & ++
\\
ξ₁ f ξ₂ & \verb!zipWith f! ξ₁ ξ₁
\\
σ f ξ & \verb!map f! ξ
\\
ξ f σ & \verb!map f! ξ
\\
$\bigtriangleup$ & \verb!take!
\\
$\bigtriangledown$ & \verb!drop!
\\
$op^{red}$ & \verb!fold op!
\\
Φ & \verb!reverse!
\\
Θ & \verb!rotate!
\\
$\varobslash$ & generalized \verb!transpose!
\\
$f Ω_d ξ$ & -- Apply $f$ along the dimension $d$ of $ξ$
\\
\hline
\end{tabular}
